D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-05-13 01intro_preliminary.tex
Update

Dirac notation~\cite{dirac1939new}, also known as bra-ket notation, is a mathematical formalism for representing quantum states using linear algebra notation. For example, Dirac notation uses the linear combination \( a\ket{\psi} + b\ket{\phi} \) to represent the superposition of the quantum states \( \ket{\psi} \) and \( \ket{\phi} \). Another essential ingredient of Dirac notation is the tensor product $\otimes$, which is used to describe composite states. For instance, the tensor expression \( \ket{\psi} \otimes \ket{\phi} \) denotes the composition of the two quantum states \( \ket{\psi} \) and \( \ket{\phi} \). A variant of Dirac notation, called labelled Dirac notation, is often used to describe composite quantum states. In labelled Dirac notation, bras and kets are tagged with labels to identify the subsystems they operate on. For example, the labelled tensor expression \( \ket{\psi}_{S'} \otimes \ket{\phi}_{S} \) indicates that \( \ket{\psi}_{S} \) and \( \ket{\phi}_{S'} \) describe two quantum states over some subsystem $S$ and $S'$, respectively. By considering the relationship between $S$ and $S'$, one can obtain identities for free, e.g.\, 
%
$$\ket{\phi}_S \otimes \ket{\psi}_{T} = \ket{\psi}_{T} \otimes \ket{\phi}_S \quad \text{if}~S \cap S' = \emptyset$$
%
In turn, commutativity of the tensor product ensures that one can
reason locally about quantum systems, and contributes to making labelled Dirac notation a convenient, compositional formalism for reasoning about quantum states, akin to how bunched logics support compositional reasoning about mutable states.

>>>>

Dirac notation~\cite{dirac1939new}, also known as bra-ket notation, is a mathematical formalism for representing quantum states using linear algebra notation. For example, Dirac notation uses the linear combination \( a\ket{\psi} + b\ket{\phi} \) to represent the superposition of the quantum states \( \ket{\psi} \) and \( \ket{\phi} \). Another essential ingredient of Dirac notation is the tensor product $\otimes$, which is used to describe composite states. For instance, the tensor expression \( \ket{\psi} \otimes \ket{\phi} \) denotes the composition of the two quantum states \( \ket{\psi} \) and \( \ket{\phi} \). A variant of Dirac notation, called labelled Dirac notation, is often used to describe composite quantum states. In labelled Dirac notation, bras and kets are tagged with labels to identify the subsystems they operate on. For example, the labelled tensor \( \ket{\psi}_{S'} \otimes \ket{\phi}_{S} \) indicates that \( \ket{\psi}_{S} \) and \( \ket{\phi}_{S'} \) describe two quantum states over subsystem $S$ and $S'$, respectively. By considering the relationship between $S$ and $S'$, one can obtain identities for free, e.g.\, 
%
$$\ket{\phi}_S \otimes \ket{\psi}_{T} = \ket{\psi}_{T} \otimes \ket{\phi}_S \quad \text{if}~S \cap S' = \emptyset$$
%
In turn, commutativity of the tensor product ensures that one can
reason locally about quantum systems, and contributes to making labelled Dirac notation a convenient, compositional formalism for reasoning about quantum states, akin to how bunched logics support compositional reasoning about mutable states.

### 2025-05-13 05tool_eval.tex
Update

\section{Implementation and Case Study}
We present \textbf{D-Hammer}, an open-source
% and publicly available\footnote{https://github.com/LucianoXu/D-Hammer}
implementation of our approach\footnote{The toll will be made publicly
available in the final version of the paper}.  \textbf{D-Hammer} is an
equational prover for Labelled Dirac notation written in \CC.  It
features a parser built using ANTLR4, and scalar reasoning is powered
by the Mathematica Engine. Users can use commands to make definitions
and assumptions in the maintained context, conduct the normalization
and equivalence checking, and obtain the rewriting trace output.
D-Hammer can be run interactively from the command line or integrated
into other \CC\ projects as a library.

>>>>

\section{Implementation and Case Study}
We present \textbf{D-Hammer}, an open-source
and publicly available\footnote{\texttt{https://github.com/LucianoXu/D-Hammer}}
implementation of our approach.  \textbf{D-Hammer} is an
equational prover for Labelled Dirac notation written in \CC.  It
features a parser built using ANTLR4, and scalar reasoning is powered
by the Mathematica Engine. Users can use commands to make definitions
and assumptions in the maintained context, conduct the normalization
and equivalence checking, and obtain the rewriting trace output.
D-Hammer can be run interactively from the command line or integrated
into other \CC\ projects as a library.

### 2025-05-13 06related_conclusion.tex
Update

\paragraph*{Comparison with DiracDec}
Xu \emph{et al}~\cite{diracdec} define a language and an associate and
commutative rewriting system for Dirac notation, and implement their
language and rewriting system in Mathematica. Our approach follows a
similar pattern. However, there are significant differences in terms
of scope, expressiveness, and efficiency. We review these differences
below.

@@

>>>>

\paragraph*{Comparison with DiracDec}
Xu \emph{et al}~\cite{diracdec} define a language and an associate and
commutative rewriting system for Dirac notation, and implement their
language and rewriting system in Mathematica. Our approach follows a
similar pattern. However, there are significant differences in terms
of scope, expressiveness, and efficiency. 
% We review these differences below.

@@

\subsection*{Acknowledgements}
\small
We thank the anonymous reviewers for their constructive feedback that helped improve this paper. We also extend our gratitude to Jam Kabeer Ali Khan and Ivan Ariel Renison for their work of testing of D-Hammer.

\normalsize

%% Disclosure of Interests
\subsubsection*{Disclosure of Interests}
\small
This research was supported by the National Key R\&D Program of China under Grant No. 2023YFA1009403.
\normalsize
